---
title: "Assemble Network"
author: "João Vitor F. Cavalcante"
date: "`r Sys.setlocale('LC_TIME', 'C'); format(Sys.time(), '%d %B, %Y')`"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../reports/") })
output:
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: false
    theme:
      bslib: true
      bootswatch: minty
---

```{r knitr, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

## PPI Network assembly from Metanalysis genes

We took all genes resulting from the metanalysis and assembled a PPI network for them by querying
STRINGdb v12.0.

Evidence channels used to gather the interactions were experimental, co-expression and databases. Said interactions were also filtered for a confidence level of 0.5.

### Load libraries and data

```{r setup}
# remotes::install_github("daniloimparato/easylayout", ref = "dadamorais")
library(easylayout)
library(tidygraph)
library(igraph)
library(ggraph)
library(dplyr)
library(tidyr)
library(vroom)
library(here)
source(here("R/stringdb.R"))

meta <- vroom(here("data/meta_analysis_genes.csv")) %>%
  select(gene_symbol, MD, pvalor, `Grouped process`)
```

### Assemble Network

```{r assemble, eval=FALSE}
string_ids <- get_string_ids(meta$gene_symbol)

meta_merged <- meta %>%
  left_join(string_ids %>% select(queryItem, stringId),
            by = c("gene_symbol" = "queryItem")) %>%
  mutate(stringId = stringr::str_remove(stringId, "9606.")) %>%
  separate_rows(`Grouped process`, sep = " \\| ")

encoded_source <- meta_merged %>%
  mutate(n = 1) %>%
  pivot_wider(
    id_cols     = stringId,
    names_from  = `Grouped process`,
    values_from = n,
    values_fn   = list(n = length),
    values_fill = list(n = 0),
    names_prefix = "From "
  ) %>%
  mutate(source_count = starts_with("From ") %>% across %>% rowSums)

network <- get_string_network(string_ids$stringId)

network_separated <-  network %>%
  separate(stringId_A,
           into = c("ncbi_taxon_id", "stringId_A"),
           sep = "\\.") %>%
  separate(stringId_B,
           into = c("ncbi_taxon_id", "stringId_B"),
           sep = "\\.")

nodelist <-
  data.frame(node = unique(c(network_separated$stringId_A, network_separated$stringId_B))) %>%
  left_join(meta_merged, by = c("node" = "stringId")) %>%
  left_join(encoded_source, by = c("node" = "stringId")) %>%
  distinct(node, gene_symbol, .keep_all = TRUE)

network_filtered <- network %>%
  combinescores(.,
                evidences = c("ascore", "escore", "dscore"),
                confLevel = 0.5) %>%
  separate(stringId_A,
           into = c("ncbi_taxon_id", "stringId_A"),
           sep = "\\.") %>%
  separate(stringId_B,
           into = c("ncbi_taxon_id", "stringId_B"),
           sep = "\\.") %>%
  dplyr::select(stringId_A, stringId_B)
```

### Compute Network Layout

```{r layout, eval=FALSE}
graph <-
  graph_from_data_frame(network_filtered, directed = FALSE, vertices = nodelist)

layout <- easylayout::vivagraph(graph)
layout <- easylayout::vivagraph(graph, layout = layout, pin_nodes = TRUE, lcc_margin_left = 10)
V(graph)$x <- layout[, 1]
V(graph)$y <- layout[, 2]

save(graph, nodelist, file = here("results/full_network.rda"))
```

### Network Plots

#### By significance

```{r}
load(here("results/full_network.rda"))

graph_colored <- graph %>%
  as_tbl_graph() %>%
  activate(nodes) %>%
  mutate(signif_md = ifelse(pvalor < 0.05, MD, NA))

ggraph(graph_colored,
       "manual",
       x = V(graph)$x,
       y = V(graph)$y) +
  geom_edge_link0(edge_width = 0.2, color = "#90909020") +
  geom_node_point(aes(color = signif_md)) +
  scale_colour_gradientn(colours = RColorBrewer::brewer.pal(9, "OrRd")) +
  coord_fixed() +
  theme_void() +
  theme(
    legend.key.size = unit(0.5, 'cm'),
    legend.key.height = unit(0.5, 'cm'),
    legend.key.width = unit(0.5, 'cm'),
    legend.title = element_text(size=6),
    legend.text = element_text(size=6),
    plot.title = element_text(size = 4, face = "bold")
  ) +
  labs(
    color = "Expression"
  )

ggsave(here("results/full_network.pdf"), width = 8, height = 8)
ggsave(here("results/full_network.png"), width = 8, height = 8, bg = "white")
```

#### By grouped process

```{r}
ggraph(graph, "manual", x = V(graph)$x, y = V(graph)$y) +
  geom_edge_link0(color = "#90909020") +
  scatterpie::geom_scatterpie(
    cols = colnames(nodelist)[startsWith(colnames(nodelist), "From ")],
    data = igraph::as_data_frame(graph, "vertices"),
    colour = NA,
    pie_scale = 0.2
  ) +
  coord_fixed() +
  theme_void() +
  theme(
    legend.key.size = unit(0.5, 'cm'),
    legend.key.height = unit(0.5, 'cm'),
    legend.key.width = unit(0.5, 'cm'),
    legend.title = element_text(size=6),
    legend.text = element_text(size=6),
    legend.position = "bottom",
    plot.title = element_text(size = 4, face = "bold")
  ) +
  labs(
    fill = "Source:"
  )

ggsave(here("results/network_processes.pdf"), width = 8, height = 8)
ggsave(here("results/network_processes.png"), width = 8, height = 8, bg = "white")
```



