---
title: "Get Related Genes"
author: "João Cavalcante"
date: "`r Sys.setlocale('LC_TIME', 'C'); format(Sys.time(), '%d %B, %Y')`"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../reports/") })
output:
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: false
    theme:
      bslib: true
      bootswatch: minty
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  eval = FALSE
)
```

# Extract Autophagy and Mitophagy related genes

```{r load-libraries}
library(readr)
library(dplyr)
library(purrr)
library(biomaRt)
library(magrittr)
library(KEGGREST)

get_go_genes <- function(go_term) {
  ensembl <- useEnsembl(biomart = "ensembl", dataset = 'hsapiens_gene_ensembl')

  getBM(
    attributes = c(
      'external_gene_name',
      'go_id',
      'name_1006'
    ),
    filters = 'go',
    values = go_term,
    mart = ensembl
  )
}
```


## Get GO genes

GO terms:

  * autophagy - [GO:0006914](https://amigo.geneontology.org/amigo/term/GO:0006914)
  * autophagy of mitocondrion - [GO:0000422](https://amigo.geneontology.org/amigo/term/GO:0000422)
  * regulation of autophagy - [GO:0010506](https://amigo.geneontology.org/amigo/term/GO:0010506)
  * negative regulation of autophagy - [GO:0010507](https://amigo.geneontology.org/amigo/term/GO:0010507)
  * positive regulation of autophagy - [GO:0000422](https://amigo.geneontology.org/amigo/term/GO:0000422)
  * mitochondrial organization - [GO:0007005](https://amigo.geneontology.org/amigo/term/GO:0007005)
  * mitochondrial fusion - [GO:0008053](https://amigo.geneontology.org/amigo/term/GO:0008053)
  * mitochondrial fission - [GO:0000266](https://amigo.geneontology.org/amigo/term/GO:0000266)

```{r get-go}
go_terms <- c(
  autophagy = "GO:0006914",
  autophagy_of_mitocondrion = "GO:0000422",
  regulation_of_autophagy = "GO:0010506",
  negative_regulation_of_autophagy = "GO:0010507",
  positive_regulation_of_autophagy = "GO:0000422",
  mitochondrial_organization = "GO:0007005",
  mitochondrial_fusion = "GO:0008053",
  mitochondrial_fission = "GO:0000266"
)

go_selected <- map_dfr(go_terms, get_go_genes)

go_renamed <- go_selected %>%
  filter(go_id %in% go_terms) %>%
  filter(external_gene_name != "") %>%
  mutate(database = "GO") %>%
  rename(gene_symbol = external_gene_name,
         annotation_id = go_id,
         annotation_name = name_1006)
```

## Get KEGG genes

KEGG Pathways / BRITE:

  * Mitochondrial Biogenesis - [br:hsa03029](https://www.kegg.jp/brite/hsa03029)
  * Autophagy - [path:hsa04140](https://www.kegg.jp/pathway/hsa04140)
  * Mitophagy - [path:hsa04137](https://www.kegg.jp/pathway/hsa04137)

We start by downloading gene to pathway mappings from the KEGG API.

```{r download}
download.file(
  "https://ftp.ncbi.nlm.nih.gov/gene/DATA/GENE_INFO/Mammalia/Homo_sapiens.gene_info.gz",
  here::here("data/Homo_sapiens.gene_info.gz")
)

download.file(
  "http://rest.kegg.jp/link/pathway/hsa",
  here::here("data/link_pathway_entrez_hsa.tsv")
)

download.file(
  "http://rest.kegg.jp/link/brite/hsa",
  here::here("data/link_brite_entrez_hsa.tsv")
)
```

Then, we get the intersection between the datasets.

```{r load}
pathways <- tibble::tribble(
  ~pathway_id,      ~pathway_name,
  "br:hsa03029",  "Mitochondrial Biogenesis",
  "path:hsa04140",  "Autophagy",
  "path:hsa04137",  "Mitophagy"
)

link_pathway_entrez_hsa <- read_tsv(
  here::here("data/link_pathway_entrez_hsa.tsv"),
  col_names = c("entrez_id", "pathway_id"),
  col_types = "cc"
) %>% tidyr::separate(
  entrez_id,
  into = c("species", "entrez_id"),
  sep = ":"
)

link_brite_entrez_hsa <- read_tsv(
  here::here("data/link_brite_entrez_hsa.tsv"),
  col_names = c("entrez_id", "pathway_id"),
  col_types = "cc"
) %>% tidyr::separate(
  entrez_id,
  into = c("species", "entrez_id"),
  sep = ":"
)

# Filtering for pathways/brite of interest
gene_pathways_hsa <- inner_join(link_pathway_entrez_hsa, pathways)
gene_brite_hsa <- inner_join(link_brite_entrez_hsa, pathways)

genes_kegg <- rbind(gene_pathways_hsa, gene_brite_hsa)
```

Finally, we extract the gene symbols.

```{r gene-symbols}
entrez_names_hsa <- read_tsv(
  here::here("data/Homo_sapiens.gene_info.gz"),
  skip = 1,
  col_names = c("entrez_id", "gene_symbol"),
  col_types = cols_only("-", "c", "c")
)

genes_kegg_hsa <- genes_kegg %>%
  inner_join(pathways) %>%
  inner_join(entrez_names_hsa) %>%
  dplyr::select(-c(species, entrez_id)) %>%
  rename(annotation_id = pathway_id,
         annotation_name = pathway_name) %>%
  mutate(database = "KEGG")
```

## Merging and writing final dataset

```{r write}
merged <- bind_rows(go_renamed, genes_kegg_hsa)

merged_collapsed <- merged %>%
  group_by(gene_symbol) %>%
  summarise(
    annotation_ids = paste0(unique(annotation_id), collapse="|"),
    annotation_names = paste0(unique(annotation_name), collapse="|"),
    databases = paste0(unique(database), collapse="|"),
    n_annotation = length(unique(annotation_id))
  )

write_csv(merged, here::here("results/selected_genes.csv"))
write_csv(merged_collapsed, here::here("results/selected_genes_collapsed.csv"))
```

